Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow

### 2024-04-17 manuscript/Manuscript.qmd
edits, refs ect

---
title: Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  Ecosystem Condition Accounts (ECA) rely on spatially representative indicators for condition. All ECAs are subject to data constraints in some way. Therefore, being able to make use of spatially biased data sets would be very valuable. For national ECAs, modelling aproaches can in some cases be used to control for sampling bias. For local ECAs however, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area. In this study we develop three ecosystem condition indicators from the same spatially biased data set from nature type mapping of Norwegian mires. These indicators are Alien species, Trenching and Antropogenic Disturbance to Soil and Vegetation (ADSV). We test our approach in three municipalities in south Norway. A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs). The spatially aggregated indicator values were calculated explicity for each HIA (ie. a stratification). To facilitate religable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately. Then we use area-weighted resampling and produce similar propabiliy distributions for each municipality. With this Bayesian updating and resampling approach, small sample sizes can be compensated for by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation. This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey. The workflow can be used at any scale. By calculating indicator values for unique spatial strata, local governemnts or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultuimately better spatial indicator coverage.
keywords: 
  - alien species
  - disturbance
  - indicator
  - ecosystem condition
  - wetlands
  - mire
  - peatlands
  - ecosystem accounting
  - SEEA EA
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

# infrastructure index
path_infrastructure <- ifelse(dir == "C:",
  "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
  "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif"
)

@@

Ecosystem condition accounting is the process of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format [@comte2022].
Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
The standard, or framework, is a set of rules, principles and best practices for compiling Ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are *variables* and *indicators*.
These are metrics chosen to reflect the central condition characteristics of the ecosystems.
These metrics are quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference condition.
Similarly, a value of 0 means that the variable is in the worst possible state, for example that the species or ecosystem function in question is completely lost from the ecosystem.
The reference condition needs to be defined for each ecosystem condition assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.

The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative sizes of the EAs [@unitednations2021 §5.54].
Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies along gradients of anthropogenic pressures and hence ecosystem condition.
This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to reliably estimate an indicator value for each EA.
This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will get its own unbiased indicator value.
There are at least 3 ways to achieve this same complete areal coverage from what is originally point-based indicator values:

a.  Predict (project) indicator values from a statistical model that accounts for the effect of environmental variation on the variables
b.  Take a central tendency (e.g. the mean) from the sample population and project it to all EAs of a given ecosystem type.
c.  Take a central tendency from a stratified sections of the sample population and projection them to the EAs, adhering to the corresponding strata.

This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].

@@

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option *a* in the list above).
Models can describe the general associations between the spatially sampled data and the context (e.g. the set of environmental variables) where it was sampled, and use these relationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can make very precice and good indicators.
This is especially true when the EAAs are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from the spatial extrapolation of such models becomes unacceptable.

@@

-   Calcareous southern fens (minerotrophic mires) \>500 m^2^

-   Calcareous northern fens \>1000 m^2^

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones [@asbjørnmoen1998].
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
We included data from 2018 to 2023.
In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is so much the case for our variables (see below).
However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable for only recorded in fens.
The other variables are recorded in all delineated mires.

From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
Variable 1, 7FA, represents the abundance of alien species, mainly plants.\
Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
These variables were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
This was done by summing the four variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
Variable 6, 7GR-GI, is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.

@@

The variable 7TK was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV*.
For Alien species and ADSV, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
For *Trenching* we used the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighbouring (ecologically speaking) mapping unit.
See Figure S1 and @fig-workflowExample for a schematic workflow for the indicator development.

According to the SEEA ECT, *Alien species* can be attributed to ECT class B1 - Compositional state characteristics and *ADSV* and *Trenching* attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].

@@

We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, Figures S3, S4).

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
|----------|----------|----------|----------|------------------------|----------|
| 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unit-less, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats laid over the area that has some sign of abration                                                              | @runehalvorsen2019   |
| 3   | PRSL                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 4   | 7TK                  | Tracks from large vehicles                       | Unit-less, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats laid over the area that has some signs of vehicle tracks                                                     | @runehalvorsen2019   |
| 5   | PRTK                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 6   | 7GR-GI               | Trenching intensity                              | Unit-less, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
| 7   | Infrastructure Index | Infrastructure Index                             | Unit-less linear scale from 0 to 13.2                                                    | Unit-less index ranging from from 0 to 13.2                                                                                                                       | @erikstad_index_2023 |
| 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                    | A categorical representation of the infrastructure index                                                                                                          | this paper           |

@@

Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
We then categorised this continuous variable into four classes (0–3) using the following value ranges: 0 = \<1; 1 ≤ 1 \> 6; 6 ≤ 2 \>12; 3 ≤ 12.
These ranges were chosen as they produced a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each range across Norway.
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.

The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.

We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set [@fig-workflowExample C; Appendix C, line XXX].
We refer to this step as the first of two *aggregation* steps (Fig. S1).
The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
This step is referred to as *spreading* the data (Fig. S1).
For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Fig. S1).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

@@

## Indicator validty and trends in Norway

The three indicators showed some association with the HIAs when we looked at data from all of Norway with `r nrow(naturetypes)` individual mires.
The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
This relationship was much stronger for the indicator Trenching, and weakest for the indicator Alien species.
Alien species also had the highest indicator values overall, with most mires having no alien species recorded.
For Trenching, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.

@@

Only in Gran municipality was there a significant difference in the Trenching indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
Conversely, the ADSV indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
The credible intervals are widest for the Trenching indicator, and this is because the national data set showed more variation in the underlying variable 7GR-GI compared to the other variables 1-5 (@tbl-variables), and this information is informing the Bayesian updating process (see @sec-methods).
Besides Nordre Follo, which do not contain any HIA-0 areas (i.e. relatively wild areas), the three municipalities had areas in all HIA classes.
Yet, none of the three municipalities had survey data from all HIAs.
Therefore, when we transferred the mean and SD for each HIA (@fig-forest) over to the ecosystem delineation map (@fig-spread-example), some mire polygons in the ecosystem delineation map did not get assigned any data, and were therefore not included (i.e. did not have any influence on) the following aggregation to ecosystem accounting areas (EAA).
At the municipality level (i.e. the EAA level) the three municipalities show some differentiation.
Nordre Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S1).

>>>>

---
title: Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  Ecosystem Condition Accounts (ECA) should reflect the integrity or quality of all nature inside the scope of the account, and therefore rely on spatially representative indicators for condition. All ECAs are subject to data constraints in some way. Therefore, being able to make use of spatially biased data sets would be very valuable. For national ECAs, modelling aproaches can in some cases be used to control for sampling biases. For local ECAs however, like at the scale of individual municipalities of developemetn projects, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area. In this study we develop three ecosystem condition indicators from the same spatially biased data set based on nature type mapping of Norwegian mires. The indicators are Alien species, Trenching and Anthropogenic Disturbance to Soil and Vegetation. We test our approach in three municipalities in south Norway. A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs). To facilitate religable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately. Then we use area-weighted resampling and produce indicator propabiliy distributions for each municipality. With this Bayesian updating and resampling approach, small sample sizes can be compensated by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation. This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey so that the data can be used for indicator in ECAs. The workflow can be used at any scale, including national scale. Because indicator values are calculatind for unique spatial strata, local governments or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultimately better spatial indicator coverage.
keywords: 
  - alien species
  - disturbance
  - ecosystem accounting
  - ecosystem condition
  - horizontal aggregation
  - indicator
  - mire
  - peatlands
  - SEEA EA
  - wetlands
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

# infrastructure index (i.e. land use intensity index)
path_infrastructure <- ifelse(dir == "C:",
  "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
  "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif"
)

@@

Ecosystem condition accounting is the process of compiling relevant data on the status, trends and qualities of ecosystems (i.e. nature) and communicating this in a structured format [@comte2022].
Its purpose is to make it easier to account for nature in policy by making the environmental costs of certain policies and practices visible to decision makers.
A statistical standard for ecosystem accounting, including ecosystem condition accounting, was developed by the UN and adopted by the UN Statistical Commission in 2021 and is called SEEA EA (@unitednations2021; System of Environmental-Economic Accounting - Ecosystem Accounting).
The standard, or framework, is a set of rules, principles and best practices for compiling ecosystem accounts, mainly aimed at national accounts.

Central to ecosystem condition accounts are *variables* and *indicators*.
These are metrics chosen to reflect the central condition characteristics of the ecosystems.
These metrics are quantified and ideally monitored over time to reflect the status and trends in condition.
Indicators are variables that are normalised (rescaled) against upper and lower reference values to become bound between the values 0 and 1.
This normalisation ensures that indicators are more comparable because an indicator value of 1 will mean the same for all indicators, i.e. that the variable equals the upper reference value which again reflect the value of the variables under the reference (pristine) condition.
Similarly, a value of 0 means that the variable is in the worst possible state, for example that the species or ecosystem function in question is completely lost from the ecosystem.
The reference condition needs to be defined for each ecosystem condition assessment separately, but SEEA EA gives some suggestions, such as an an ecosystem with no or minimal anthropogenic disturbance.

The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the EAs [@unitednations2021 §5.54].
Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to reliably estimate an indicator value for each EA.
This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will have ebough data to estaimte its unique indicator value.
This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
There are at least three ways to achieve this same complete areal coverage from indicator values that originally are spatially incomplete:

a.  Predict (project) indicator values from a statistical model that accounts for the effect of environmental variation on the variables
b.  Take a central tendency (e.g. the mean) from the sample population and project it to all EAs of a given ecosystem type.
c.  Take a central tendency from stratified sections of the sample population and projection them to the EAs, adhering to the corresponding strata.

@@

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option *a* in the list above).
Statistical models can describe the general associations between the sampled data and the context where it was sampled (e.g. a set of environmental variables), and use these relationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can make very precice and good indicators.
This is especially true when the EAAs are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from the spatial extrapolation of such models becomes unacceptable.

@@

-   Calcareous southern fens (minerotrophic mires) \> 500 m^2^

-   Calcareous northern fens \> 1000 m^2^

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones [@asbjørnmoen1998].
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
We included data from 2018 to 2023.
In this paper we assume the survey is representative for the entire mire ecosystem in Norway (i.e. that there is no thematic bias).
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed, we do not think this is so much the case for our variables (see below).
However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable for only recorded in fens.
The other variables are recorded in all delineated mires.

From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
Variable 1, 7FA, represents the abundance of alien species, mainly plants.
Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
These variables were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
This was done by summing the four variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
Variable 6, 7GR-GI, is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.

@@

The variable 7TK was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV* (Figure S1).
For *Alien species* and *ADSV*, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
For *Trenching* we used the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighbouring (ecologically speaking) mapping unit.
See @fig-workflowExample and S1 for schematic workflows for the indicator development.
*Alien species* was attributed to Ecosystem Condition Typology (ECT) class B1 - Compositional state characteristics and *ADSV,* and *Trenching* was attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].

@@

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                         | Description                                                                                                                                                       | Reference            |
|------------|------------|------------|------------|------------|------------|
| 1   | 7FA                  | Prevalence of alien species                      | Unit-less, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019   |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unit-less, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats laid over the area that has some sign of abration                                                              | @runehalvorsen2019   |
| 3   | PRSL                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 4   | 7TK                  | Tracks from large vehicles                       | Unit-less, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats laid over the area that has some signs of vehicle tracks                                                     | @runehalvorsen2019   |
| 5   | PRTK                 | *as above*                                       | Unit-less, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet   |
| 6   | 7GR-GI               | Trenching intensity                              | Unit-less, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet   |
| 7   | Infrastructure Index | Infrastructure Index                             | Unit-less linear scale from 0 to 13.2                                                    | Unit-less index ranging from from 0 to 13.2                                                                                                                       | @erikstad_index_2023 |
| 8   | HIA                  | Homogeneous Impact Area                          | Ordinal, non-linear scale from 1 to 4                                                    | A categorical representation of the infrastructure index                                                                                                          | this paper           |

@@

Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
We then categorised this continuous variable into four classes (0–3) using ranges chosen to produce a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each strata across Norway (Table S1).
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.
The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.

We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, S3, S4).
We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set (@fig-workflowExample C; Appendix C, line 138 <!--# update line number manually when re-rendering appendix C -->).
We refer to this step as the first of two *aggregation* steps (Figure S1).
The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
This step is referred to as *spreading* the data (Figure S1).
For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Figure S1).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

All the data preparation and analyses were done in RStudio [@positteam2024] and R version 4.3.0 [@rcoreteam2023], relying heavily on the packages tidyverse [@wickham2019] , sf [@pebesma2018] and stars [@pebesma2023].

@@

## Indicator validity and trends in Norway

The three indicators showed some association with the HIAs when we looked at data from all of Norway with `r nrow(naturetypes)` individual mires.
The indicators mostly showed a worsening of condition with increasing presence of human infrastructure (@fig-validation).
This relationship was much stronger for the indicator *Trenching*, and weakest for the indicator *Alien species*.
*Alien species* also had the highest indicator values overall, with most mires having no alien species recorded.
For *Trenching*, 57% of mires in HIA-3 had some trenches, whereas in HIA-0 this number was 10%.

@@

Only in Gran municipality was there a significant difference in the *Trenching* indicator between HIAs, with HIA-2 having worse condition than HIA-1 (@fig-forest).
Conversely, the *ADSV* indicator in Gran showed worse condition in HIA-1 compared to HIA-2.
Nordre-Follo had considerably fewer data points compared to Gran, and especially to Nord-Aurdal who had the most data points, and this paucity of data is reflected in the wide credible intervals for all three indicators in Nordre Follo.
The credible intervals are widest for the *Trenching* indicator, and this is because the national data set showed more variation in the underlying variable 7GR-GI compared to the other variables 1-5 (@tbl-variables), and this information is informing the Bayesian updating process (see @sec-methods).
Besides Nordre Follo, which do not contain any HIA-0 areas (i.e. relatively wild areas), the three municipalities had areas in all HIA classes.
Yet, none of the three municipalities had survey data from all HIAs.
Therefore, when we transferred the mean and SD for each HIA (@fig-forest) over to the ecosystem delineation map (@fig-spread-example), some mire polygons in the ecosystem delineation map did not get assigned any data, and were therefore not included (i.e. did not have any influence on) the following aggregation to ecosystem accounting areas (EAA) (@fig-workflowExample, E) .
At the municipality level (i.e. the EAA level) the three municipalities show some differentiation.
Nordre Follo had the highest (the best) indicator values for ADSV, but the lowest for Trenching, which was also the only instance of an indicator crossing the threshold from good to deteriorated condition (@fig-EEA-ridgePlot; Table S2).

```{r fig-workflowExample}
#| eval: true
#| include: true
#| out.width: '100%'
#| fig.cap:
#| - "Simplified schematic showing the workflow (Panes A-E) for spatial generalisation of a spatially biased ecosystem condition indicator. Pane A shows a spatially explicit, patchy and spatially biased indicator. The outline is the ecosystem accounting area (EEA). Pane B shows the location of four homogeneous impact areas (HIAs) inside the EEA. The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area woeghted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)). The bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set. The width of the posterior distribution responds both to variation in the data and to the sample size, giving  a realistic measure for the uncertainty around the indicator even with a singe observation. The distributions in Pane C are assumed normally distributed. Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2. In Pane D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately. The colors indicator different indicator values (highlighted in the inset). For each ecosystem occurence (i.e. ecosystem assets) in Pane D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon. In Pane E we have randomly sampled 1000 values from the entire vector of samples in the previous step, and we get an area weighted probability distribution for the mean indicator value for the EEA."
